# Introduction to R for Absolute Beginners

Welcome to R! This guide is designed to help absolute beginners get started with R, a powerful tool for statistical computing and data visualization. R can seem overwhelming at first, but this introduction will walk you through the basics, step by step. If you are already familiar with R, feel free to skip this section.

## What is R?

R is a programming language and software environment specifically designed for statistical analysis and data visualization. It is widely used by data scientists, statisticians, and researchers.

## Getting Started with R

### Installing R and RStudio

To use R, you will need two things:
1. **R**: Download and install it from [CRAN](https://cran.r-project.org/).
2. **RStudio**: A user-friendly interface for R, which you can download from [RStudio's website](https://posit.co/download/rstudio-desktop/).

### RStudio Interface

When you open RStudio, you'll see four main panels:
- **Source/Script Editor**: Write and save your R code here. To create an R script go to *File -> New File -> R Script*.

- **Console**: Run code directly and see the output.

- **Environment/History**: See your data and objects.

- **Files/Plots/Packages/Help**: Access files (available in your working directory, see below), view plots, manage packages, and get help. you can easily load data sets using this option.


## Working Directory

The working directory is the folder where R will look for files. You can set your working directory using the `setwd()` function. Alternatively, you can use the `here` package to set your working directory to the location of your R script. You can also navigate on top of R studio to the session tab and set your working directory from there. Finally you can use the keyboard shortcut `Ctrl + Shift + H` to set your working directory. If you cannot load a certain file (usually a data set), it is likely that your working directory is not set correctly.


## Basic Syntax in R

### Running Code

You can run code directly in the console or write it in the script editor and press `Ctrl + Enter` (or `Cmd + Enter` on a Mac) to run it. We always suggest you do the later, that way your code is saved. 

### Basic Calculations

At its simplest, R can be used as a calculator. For example:

```{r, eval=FALSE}
1 + 1
5 * 3
10 / 2
2^3
```

### Variables 

You can store values in **variables** or as they sometimes called **objects**. A variable is like a box that holds data.

```{r, eval=FALSE}
x <- 5
y <- 10
sum <- x + y
sum
```

### Data Types

R has several basic data types:

- Numeric: Numbers like 42 or 3.14.
- Character: Text or strings like "Hello".
- Logical: `TRUE` or `FALSE`.

```{r, eval=FALSE}
num_var <- 42  # numeric
char_var <- "Hello!"  # string
log_var <- TRUE # logical 
```


### Functions

A function is a piece of code that performs a specific task. For example, `sqrt()` calculates the square root:

```{r, eval=FALSE}
sqrt(16)
```

There are many in-built functions in R, and you can also create your own functions (however, that is beyond the scope of this book). And of course, as you will see in the other chapters, many of the functionalities you will use come from packages, which can be installed and loaded follows:

```{r, eval=FALSE}
install.packages("package_name")
library(package_name)
```

## Working with Data & Various Data Structures 

### Vectors

A vector is a sequence of data elements of the same basic type. You can create a vector using the `c()` function:

```{r, eval=FALSE}
my_vector <- c(1, 2, 3, 4, 5)
my_vector
```


### Matrices

A matrix is a two-dimensional data structure where all elements are of the same type (usually numeric). Unlike a data frame, a matrix does not allow for different types of data (e.g., characters and numbers) in different columns. You can create a matrix using the `matrix()` function:

```{r, eval=TRUE}
my_matrix <- matrix(
  data = 1:9,      # The data to fill the matrix
  nrow = 3,        # Number of rows
  ncol = 3,        # Number of columns
  byrow = TRUE     # Fill the matrix by rows
)
my_matrix
```

The concept of a matrix is closely tied to network models, since in practice the estimated network model is a matrix of connections between nodes. Below is an example of constructing a symmetric 5x5 adjacency matrix (i.e., a a graph that shows which nodes are present without carrying information about the strength of the nodes, zeros indicate absent edges and ones present edges) for an undirected graph and visualizing it using the qgraph package:

```{r, warning=FALSE, message=FALSE, eval=TRUE}
# Create a 5x5 symmetric adjacency matrix
adj_matrix <- matrix(
  data = c(0, 1, 1, 0, 1,
           1, 0, 1, 1, 0,
           1, 1, 0, 1, 0,
           0, 1, 1, 0, 1,
           1, 0, 0, 1, 0),
  nrow = 5,
  ncol = 5,
  byrow = TRUE
)

# Add row and column names for clarity
colnames(adj_matrix) <- rownames(adj_matrix) <- c("Node1", "Node2", "Node3", "Node4", "Node5")

# Print the adjacency matrix
adj_matrix

# Plot the graph using qgraph
library(qgraph)
qgraph(adj_matrix, 
       layout = "spring", 
       vsize = 8, 
       labels = colnames(adj_matrix))
```

Notice that adjacency matrices are a specific use of matrices where symmetry often indicates undirected relationships, and zeros along the diagonal reflect the absence of self-loops.

### Data Frames

A data frame is a table-like data structure (i.e., a spreadsheet) that stores data. You can create a data frame using the `data.frame()` function:

```{r, eval=TRUE}
my_data <- data.frame(
  Name = c("A", "B", "C"),
  Age = c(25, 30, 35),
  Height = c(175, 182, 168)
)
my_data
```

### Matrix vs. Data Frame

While matrices and data frames might look similar at first glance (both are rectangular), there are key differences:

- Data Frames can store different types of data in each column (e.g., numeric, character, logical).

- Matrices store only one type of data across all elements.

If you convert a matrix to a data frame, it gains the flexibility of mixed data types but may lose some structure if row and column names aren't defined.


### Lists

A list in R is a flexible data structure that can hold objects of different types and sizes. Unlike a matrix or data frame, lists are not restricted to a rectangular shape. You can think of a list as a container for various objects, including vectors, matrices, data frames, or even other lists. Usually the results of many of the analyses are stored in lists as they can hold many different types of objects.

Create a list using the `list()` function:

```{r, eval=TRUE}
my_list <- list(
  Name = "A",
  Age = 25,
  Scores = c(90, 85, 88),
  Nested = list(a = 1, b = 2)
)
my_list
``` 


### Arrays

An array is a multi-dimensional generalization of a matrix. While matrices are limited to two dimensions (rows and columns), arrays can have more dimensions. Arrays store data of the same type, and you can create one using the array() function:

```{r}
my_array <- array(
  data = 1:24, 
  dim = c(4, 3, 2) # 4 rows, 3 columns, 2 layers
)
my_array
```

## Relationship to Lists and Arrays 

- Lists are more flexible than matrices and data frames, as they can store mixed types and structures. 

- Arrays are similar to matrices but can expand into more than two dimensions, making them suitable for complex datasets like 3D models or time-series grids.

## Loading Data into R

R can read data from various file formats, as already mentioned, you can also use the file tab for this. Here are some examples of how this can be done through the console, for some common file formats. Before we start, make sure to set your working directory to the location where your data is stored. 

### Reading a CSV file

```{r, eval=FALSE}
# Make sure the 'data.csv' file is in your working directory
my_data <- read.csv("data.csv")
head(my_data)
```

### Reading an Excel File

```{r, eval=FALSE}
# Install the readxl package if you haven't already
#install.packages("readxl")
library(readxl)

# Reading an Excel file
my_data <- read_excel("data.xlsx", sheet = 1)
head(my_data)
```

### Reading Data from R's Built-in Datasets

Many R packages come with built-in datasets that you can use to practice. For example, the `mtcars` dataset is built into R:


```{r, eval=TRUE}
# Load the 'mtcars' dataset
data(mtcars)
head(mtcars)
```

## Performing a statistical analysis

R is a powerful tool for statistical analysis. You can perform a wide range of statistical tests and analyses using R. Here is an example of how to perform a simple t-test.

### T-Test

```{r, eval=TRUE}
# Subset the data into two groups
auto <- mtcars$mpg[mtcars$am == 0]  # Automatic cars
manual <- mtcars$mpg[mtcars$am == 1]  # Manual cars

# Perform a t-test
t_test_result <- t.test(auto, manual)
t_test_result
```


### Visualizing data

R is also great for data visualization. You can create a wide range of plots and graphs using R. Here is an example of how to create a boxlot. R has in built functions for many types of plots, and you can also use the `ggplot2` package for more advanced plots.

```{r, eval=TRUE}
boxplot(mpg ~ am, data = mtcars,
        main = "Miles Per Gallon (mpg) by Transmission Type",
        xlab = "Transmission (0 = Automatic, 1 = Manual)",
        ylab = "Miles Per Gallon (mpg)")
```


## Additional Resources

This introduction is just the tip of the iceberg. R is a vast and powerful tool with a steep learning curve. Here are some additional resources to help you continue your journey with R:

- [R for Data Science](https://r4ds.had.co.nz/): A comprehensive guide to data science with R.
- [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/): Handy cheatsheets for various R topics.
- [Stack Overflow](https://stackoverflow.com/): A great place to ask questions and find answers about R.
- [R Documentation](https://www.rdocumentation.org/): Official documentation for R packages.


The second chapter in the book *Network Psychometrics with R* [@NP_with_R_book] also provides a comprehensive introduction to R. 

## Conclusion

Congratulations! You've completed the introduction to R. We hope this guide has given you a solid foundation to start exploring R on your own. In the following chapters, we explore the specific functionalities of the R packages that will be used throughout this workshop. If you have any questions or need further assistance, don't hesitate to ask. Happy "networking"!


