# Causality, Conditional Dependence, and Pairwise Markov Random Fields

## Overview

This chapter guides you through performing network discovery and causal analysis in R. 
For more details, please consult the other materials provided for this lecture.

- Load the necessary packages and data.
- Compute and interpret correlation and partial correlation matrices.
- Identify significant connections using the SIN algorithm.
- Visualize results with the `qgraph` package.
- Apply the `mgm` package for network estimation.

## Setup

Start by loading the required packages:

```{r setup, message=FALSE, warning=FALSE}
#install.packages("gtools")
#install.packages("corpcor")
#install.packages("qgraph")
#install.packages("mgm")
#install.packages("igraph")
#install.packages("RBGL")
#install.packages("graph")
#install.packages("pcalg")

library(gtools)
library(corpcor)
library(qgraph)
library(mgm)
library(igraph)
library(RBGL)
library(graph)
library(pcalg)
```

## Loading the Data

We will use a simulated dataset for this analysis. 

```{r load-data}
# Simulating correlated data using a multivariate normal distribution
library(MASS)

# Define the mean vector and covariance matrix
mu <- rep(0, 5) # Mean of 0 for each variable
Sigma <- matrix(0.5, nrow = 5, ncol = 5) # 0.5 correlation between variables
diag(Sigma) <- 1 # Variance of 1 for each variable

# Simulate 100 observations
set.seed(123) # For reproducibility
data <- as.data.frame(mvrnorm(n = 100, mu = mu, Sigma = Sigma))
names(data) <- paste0("V", 1:5) # Naming the variables

```


## Correlation and Partial Correlation

### Correlation Matrix

Compute the correlation matrix of the dataset:

```{r correlation}
data.cor <- cor(data)
print(data.cor)
```

### Partial Correlation Matrix

Convert the correlation matrix to a partial correlation matrix:

```{r partial-correlation}
data.pcor <- cor2pcor(data.cor)
print(data.pcor)
```


## Identifying Significant Connections


We can use a method called SIN (Significant, Intermediate, Non-significant, @pearl1995), to test which partial correlations are significantly different from zero. For that you can use the following function:

```{r}
sin.ag <-
function (pcor, n, plot = TRUE, alpha = 0.1, beta = 0.5) 
{
    p <- dim(pcor)[2]
	pval.pcor <- function(pcor, n){
		pcor.unic <- pcor[lower.tri(pcor, diag=FALSE)]
		z <- 0.5*log((1+pcor.unic)/(1-pcor.unic))
		pval <- 2*(1-pnorm(sqrt(n-3)*abs(z)))
		return(p.adjust(pval, method="holm"))
	}   
	sin.lt <- pval.pcor(pcor, n)
	sin.amat <- matrix(0,ncol=p,nrow=p)
	sin.amat[lower.tri(sin.amat)] <- sin.lt
	sin.amat <- sin.amat + t(sin.amat)
	sin.p <- sin.amat
	sin.amat[sin.p < 0.1] <- 1
	sin.amat[sin.p >= 0.1] <- 0
	diag(sin.amat) <- 0
	if (plot) {
    	connect <- combinations(p, 2)
    	lc <- dim(connect)[1]
    	make.name <- function(a) paste(a[1],paste("-",a[2],sep=""),sep="")
    	leg <- apply(connect,1,make.name)
        plot(sin.lt, pch = 16, bty = "n", axes = FALSE, xlab = "edge", 
            ylab = "P-value")
        axis(1, at = 1:lc, labels = leg, las = 2)
        axis(2, at = c(0.2, 0.4, 0.6, 0.8, 1), labels = TRUE)
        lines(c(1, lc), c(alpha, alpha), col = "gray")
        lines(c(1, lc), c(beta, beta), col = "gray")
        name <- deparse(substitute(data))
        title(main = paste(attr(data, "cond")), font.main = 1)
        text(1,0.15,"0.1",col="gray")
        text(1,0.55,"0.5",col="gray")        
    }
    names(sin.lt) <- leg
    invisible(list(pval=sin.lt,amat=sin.amat))
}
```


```{r sin-algorithm}
sin <- sin.ag(data.pcor,n=100)
sin
```


## Visualizing the Network

We can visualize the significant connections using `qgraph`:

```{r visualize-network}
qgraph(sin$amat)
```

## Network Estimation with MGM

Estimate a graphical model using the `mgm` package. For this the data needs to be saved as a matrix.

```{r mgm-analysis, message=FALSE, warning=FALSE}
fit <- mgm(as.matrix(data), type = rep("g", 5), lambdaSeq = 0, lambdaSel = "EBIC")
```



## Summary

This tutorial demonstrated how to:

- Compute and interpret correlation and partial correlation matrices.
- Identify significant connections using statistical tests.
- Visualize the network structure.
- Estimate a network model with the `mgm` package.

Explore further by doing the practical! 